Using Linear Regression to Examine the Correlation Between Access to Drinking Water and Life Expectancy

### 2022-12-09 Manuscript/Paper.tex
additions to paper

Living in the United States, we often take the necessity of safe drinking water for granted, but it is still a pressing issue in many parts of the world. Current research shows that unsafe water is responsible for millions of deaths \citep{ritchieroser2019water}. From the \textit{World Health Organization}, "Contaminated water and poor sanitation are linked to transmission of diseases such as cholera, diarrhoea, dysentery, hepatitis A, typhoid and polio" \citep{who2017water}. These diseases are largely preventable, and an increase in both access to and quality of safe drinking water can dramatically increase life expectancy \citep{angelakis2021quality}.

Although there are numerous studies on how drinking contaminated water can have tremendous negative effects on one's health, there are very few studies that examine how the percentage of people with access to clean water in a certain area affects statistics regarding the community's overall health. Previous studies regarding drinking water and health often use the presence of clean water as a binary outcome, where a community either does or does not have access. Other times, facilities lie on a continuum. In \citet{wolf2014sanitationondisease}, the quality of drinking water was measured by water and sanitation facilities that lie on a scale between poor and good. Conclusions were drawn based on the improvements to water systems, stating that "...there are large potential reductions in diarrhoeal disease risk through improvements to both water and sanitation in low- and middle-income settings" \citep{wolf2014sanitationondisease}.

Additionally, studies regarding drinking water have been done to draw conclusions about the presence of certain diseases when clean drinking water is limited. Yet, there are other external effects of maintaining access to clean drinking water. From \citet{2020mappinggeo}, "Low access to safe water and sanitation has also been linked to broader social outcomes such as reductions in school attendance (particularly for girls who are menstruating), losses to economic productivity, and undue burden on women of time spent collecting water". This study aims to quantify how access to drinking water affects a population's overall life expectancy at birth. Such a study could help emphasize the need for clean water worldwide, as it has numerous health benefits beyond longevity \citep{popkin2010waterhealth}.

@@

The two data sets we will be using are from the World Health Organization and provided by Kaggle. The first is called Basic Drinking Water Services and contains information regarding basic drinking water services across various countries and in different years. There are 3455 rows with 4 columns: Location, Period, Indicator, and Tooltip. The indicator column is a text cell that says "Population using at least basic drinking-water services", and the Tooltip column gives the percentage of the population where the indicator is true. In this data set, we will be looking at the country, year, and percentage of population using at least basic drinking water services.

The second data set is called Life Expectancy at Birth and contains information about life expectancy in various countries at the time of birth. It has 2197 rows and 5 columns: Location, Period, Indicator, Dim1, and Tooltip. In this data set, the indicator is the life expectancy at birth in years. The Dim1 column indicates which sex the life expectancy is referring to - male, female, or both sexes. For this data set, we will have to filter the rows to just contain those with a life expectancy for both sexes, as the other data set is not filtered by gender. We will be using the country and year columns to look at life expectancy.

@@

These lists will also be converted into numpy arrays.
\begin{lstlisting}[language=Python]
    DW2015=np.array(dw2015)
    lifeExpectancy2015=np.array(le2015)
\end{lstlisting}

@@

Now that we have standardized data across the board, we can plot the life expectancy and basic drinking water. The output is shown in Figure ~\ref{fig3}.
\begin{figure}
    \centering
    \includegraphics[width=4in]{Figures/figure1.png}
    \caption{Life Expectancy vs. Drinking Water}
    \label{fig3}
\end{figure}

Now we create our data matrix $x$, which is the $x$ array transformed into a column vector and appended to a column of ones.
\begin{lstlisting}[language=Python]
    x2015=x2015.reshape(-1,1)
    X2015=np.concatenate([x2015,np.ones(shape=(x2015.shape[0],1))],axis=1)
\end{lstlisting}

@@

As seen in Figure ~\ref{fig3}, we can assume a linear regression is appropriate for the model based on the data. There looks to be a linear correlation between the variables rather than a quadratic or other polynomial relationship. To see the R-Squared value of our linear regression, we can use the following code \citep{pythonr2}:
\begin{lstlisting}[language=Python]
    corr_matrix = np.corrcoef(Y2015[:,0], Yhat2015[:,0])
    corr = corr_matrix[0,1]
    R_sq = corr**2
\end{lstlisting}
We see the $R^2$ value is 0.66348. So, 66.348\% of the variability in the data is explained by the model. While this is on the lower end, our model is still supported. We would also like to check that the mean of the residuals is close to zero. This can be calculated with the following code.
\begin{lstlisting}[language=Python]
    residuals=Yhat2015[:,0]-Y2015[:,0]
    np.mean(residuals)
\end{lstlisting}
We see the mean is -1.2813065726821479e-14 which is very close to zero. We also need to check for homosteadasticity among the residuals. To look at the residuals, the difference between the measured and predicted values, we can use the following function as detailed in \citet{teitelbaum2021linreg}.
\begin{lstlisting}[language=Python]
    def comparison_plot(x,Y, Yhat):
        '''Plots Predicted vs True values for analysis of regression'''
        comparison_plot = 
            figure(title='Difference between Measured and Predicted Values')
        comparison_plot.xaxis.axis_label='x'
        comparison_plot.yaxis.axis_label='Yhat-Y'
        comparison_plot.scatter(x=x,y=Yhat-Y)
        comparison_plot.line(x=[x.min(),x.max()],y=[0,0])
        return comparison_plot
    show(comparison_plot(x2015.ravel(),Y2015[:,0], Yhat2015[:,0]))
\end{lstlisting}

@@

From Figure 1, we see a general positive correlation between the two variables. This relationship illustrates that as the percentage of the population's access to safe drinking water increases, the average life expectancy also increases. Based on prior research, like that in \citet{angelakis2021quality} that suggests access to and quality of drinking water can increase an individual's life expectancy, this result was expected. Later on, we discovered the slope of the best fit line to be 0.279, and the $R^2$ value is 0.6638. This further supports our prediction that there would be a positive relationship between access to drinking water and life expectancy.

From Figure 2 with the resulting best fit line, we see that some points are very close to the line while others are very far. Checking the plot of the residuals in Figure 3, we see there is a somewhat large variation between the predicted values and the measured values. Some of the values are extremely close to zero, meaning that there was little difference between the predicted value and the actual value. However, some of the values are closer to 10-15 away from zero, indicating a large difference between the predicted and measured values.

@@

These results highlight the importance for access to clean drinking water, as we saw a positive correlation between access to safe water and life expectancy. We also saw some variation in the residuals, meaning that the model predicted a different outcome than the data points observed. The level of variation is a discrepancy between the predicted life expectancy of the model and the actual observed life expectancy from the data. This most likely has to do with the varying factors among different countries including other health factors such as air quality, sanitation, living conditions, and political stability. More conclusive research would have to be done to verify this. Possible methods such as multivariate regression and an analysis of covariance could be done. It is known that a person's health is affected by many factors, and a complete study to analyze just how much effect each factor has would be extensive. Though a study of this scale is improbable for this paper, such findings could revolutionize what public health organizations, governments, and citizens prioritize in relation to their health. This type of study has far-reaching effects and could not only improve a person's life expectancy but also their quality of life.

Studies regarding drinking water have been done to draw conclusions about the presence of certain diseases when clean drinking water is limited. Research revolves around specific diseases or access to clean drinking water based on a scale. This research paper focuses on investigating the effect of the proportion of a population's access to clean drinking water on their life expectancy - a more widespread conclusion than those regarding specific water-borne illnesses. We have found a positive correlation between the variables, which supports existing research that clean water is essential for the body to function, yet expands the relationship to overall life expectancy. Insights in this field can further promote the basic right to safe water and highlight just how important water is in human health.

>>>>

Many of us in the United States often take the necessity of safe drinking water for granted, but there are parts of the country that still struggle with availability of clean water. In fact, "in 2015, nearly 21 million people relied on community water systems that violated health-based quality standards" \citep{2018nattrends}. Furthermore, it is still a pressing issue in many parts of the world. Current research shows that unsafe water is responsible for millions of deaths \citep{ritchieroser2019water}, and from the \textit{World Health Organization}, "Contaminated water and poor sanitation are linked to transmission of diseases such as cholera, diarrhoea, dysentery, hepatitis A, typhoid and polio" \citep{who2017water}. These diseases are largely preventable, and an increase in both access to and quality of safe drinking water can dramatically increase life expectancy \citep{angelakis2021quality}.

Although there are numerous studies on how drinking contaminated water can have tremendous negative effects on one's health, there are very few studies that examine how the percentage of people with access to clean water in a certain area affects statistics regarding the community's overall health. In previous studies, the quality of drinking water and sanitation facilities lie on a continuum. In \citet{wolf2014sanitationondisease}, the quality of drinking water was measured by water and sanitation facilities on a scale from poor to good. Conclusions were drawn based on the improvements to water systems, rather than the amount of people with access to potable water, stating that "...there are large potential reductions in diarrhoeal disease risk through improvements to both water and sanitation in low- and middle-income settings" \citep{wolf2014sanitationondisease}. Moreover, this study was based on the specific health implications regarding diarrhoeal disease risk.

In fact, studies regarding drinking water have been done to draw conclusions about the presence of certain diseases, like the study above, when clean drinking water is limited. Yet, there are many other external effects of maintaining access to clean drinking water. From \citet{2020mappinggeo}, "Low access to safe water and sanitation has also been linked to broader social outcomes such as reductions in school attendance (particularly for girls who are menstruating), losses to economic productivity, and undue burden on women of time spent collecting water". Additionally, the distance the source of the water is from one's home is extremely important. From the work done in \citet{2018waterandhealth}, "Across all countries in this study, results show that up to 40\% of the national population needs more than 30 minutes to fetch water irrespective of the type of water source used by the household...as reducing the collection burden required to fetch water is essential to enhancing water quantity available for households and improving general health and quality of life". Besides the poor health effects, there are various other negative effects that are present alongside low access to potable water. Chronic stress in one's life whether related to nutrition, health, financial well-being, or other factors is linked to a reduction in life expectancy. This study aims to quantify how access to drinking water (and the negative mental and physical health effects associated with it) affects a population's overall life expectancy at birth. Such a study could help emphasize the need for accessible clean water worldwide, as it has numerous health benefits beyond longevity \citep{popkin2010waterhealth}.

@@

The two data sets we will be using are from the World Health Organization and provided by Kaggle under the name "World Health Statistics 2020|Complete|Geo-Analysis" \citep{dataset}. The first is called Basic Drinking Water Services and contains information regarding basic drinking water services across various countries and in different years. From WHO, basic drinking water services is defined as "drinking water from an improved source, provided collection time is not more than 30 minutes for a round trip. Improved water sources include piped water, boreholes or tubewells, protected dug wells, protected springs, and packaged or delivered water" \citep{twd2022glossary}. There are 3455 rows with 4 columns: Location, Period, Indicator, and Tooltip. The indicator column is a text cell that says "Population using at least basic drinking-water services", and the Tooltip column gives the percentage of the population where the indicator is true. In this data set, we will be looking at the country, year, and percentage of population using at least basic drinking water services.

The second data set is called Life Expectancy at Birth and contains information about life expectancy in years for various countries at the time of birth. It has 2197 rows and 5 columns: Location, Period, Indicator, Dim1, and Tooltip. In this data set, the indicator is the life expectancy at birth in years. The Dim1 column indicates which sex the life expectancy is referring to - male, female, or both sexes. For this data set, we will have to filter the rows to just contain those with a life expectancy for both sexes, as the other data set is not filtered by gender. We will be using the country and year columns to look at life expectancy.

@@

These lists must also be converted into numpy arrays.
\begin{lstlisting}[language=Python]
    DW2015=np.array(dw2015)
    lifeExpectancy2015=np.array(le2015)
\end{lstlisting}

@@

Now that we have standardized data across the board, we can plot the life expectancy in years and the percentage of the population with access to basic drinking water services. The output is shown in Figure ~\ref{fig3}.
\begin{figure}
    \centering
    \includegraphics[width=4in]{Figures/figure1.png}
    \caption{Life Expectancy vs. Drinking Water}
    \label{fig3}
\end{figure}

Now we create our data matrix $x$, which is the $x$ array transformed into a column vector and appended to a column of ones.
\begin{lstlisting}[language=Python]
    x2015=x2015.reshape(-1,1)
    X2015=np.concatenate([x2015,np.ones(shape=
        (x2015.shape[0],1))],axis=1)
\end{lstlisting}

@@

As seen in Figure ~\ref{fig3}, we can assume a linear regression is appropriate for the model based on the data. There looks to be a linear correlation between the variables rather than a quadratic or other non-linear relationship. To see the R-Squared value of our linear regression, we can use the following code \citep{pythonr2}:
\begin{lstlisting}[language=Python]
    corr_matrix = np.corrcoef(Y2015[:,0], Yhat2015[:,0])
    corr = corr_matrix[0,1]
    R_sq = corr**2
\end{lstlisting}
We see the $R^2$ value is 0.66348. So, 66.348\% of the variability in the data is explained by the model. While this is on the lower end, our model is still supported. We would also like to check that the mean of the residuals is close to zero. This can be calculated with the following code.
\begin{lstlisting}[language=Python]
    residuals=Yhat2015[:,0]-Y2015[:,0]
    np.mean(residuals)
\end{lstlisting}
We see the mean is -1.2813065726821479e-14 which is very close to zero. Next, we need to check for homosteadasticity among the residuals. To look at the residuals, the difference between the measured and predicted values, we can use the following function as detailed in \citet{teitelbaum2021linreg}.
\begin{lstlisting}[language=Python]
    def comparison_plot(x,Y, Yhat):
        '''Plots Predicted vs True values for analysis of regression'''
        comparison_plot = 
            figure(title='Difference between Measured and Predicted Values')
        comparison_plot.xaxis.axis_label='x'
        comparison_plot.yaxis.axis_label='Yhat-Y'
        comparison_plot.scatter(x=x,y=Yhat-Y)
        comparison_plot.line(x=[x.min(),x.max()],y=[0,0])
        return comparison_plot
    show(comparison_plot(x2015.ravel(),Y2015[:,0], Yhat2015[:,0]))
\end{lstlisting}

@@

From Figure ~\ref{fig3}, we see a general positive correlation between the two variables. This relationship illustrates that as the percentage of the population's access to safe drinking water increases, the average life expectancy also increases. Based on prior research, like that in \citet{angelakis2021quality} that suggests access to and quality of drinking water can increase an individual's life expectancy, this result was expected. Later on, we discovered the slope of the best fit line to be 0.279, and the $R^2$ value is 0.6638. This further supports our prediction that there would be a positive relationship between access to drinking water and life expectancy.

From Figure ~\ref{fig4} containing the resulting best fit line, we see that some points are very close to the line while others are very far. Checking the plot of the residuals in Figure ~\ref{fig5}, we see there is a somewhat large variation between the predicted values and the measured values. Some of the values are extremely close to zero, meaning that there was little difference between the predicted value and the actual value. However, some of the values are closer to 10-15 away from zero, indicating a large difference between the predicted and measured values. This confirms the observation made from Figure ~\ref{fig4}.

@@

Once the data was adjusted for correlating countries in the same year, the scatterplot suggested a positive linear relationship existed. After the line of best fit was computed, this was confirmed. The slope of the least squares line is 0.279 with an intercept of 49.157. Theoretically, if $0\%$ of the population had access to clean drinking water, we would expect the average life expectancy at birth to be 49.157. Every one-unit increase in the percentage of access to drinking water results in a 0.279 year increase on average in the average life expectancy of the population. These results highlight the importance for access to basic drinking water services, as we saw a positive correlation between the percentage of the population with access to these basic services and life expectancy at birth.

\section{Limitations}

This study was only done for the year 2015. While it was necessary to keep the year constant, it would be beneficial to research other years. It would be interesting to see if the correlation stayed somewhat constant or varied among the years. Performing the regression on multiple years would also help validate our results. Additionally, we saw some variation in the residuals, meaning that the model predicted a different outcome than the data points observed. The level of variation is a discrepancy between the predicted life expectancy of the model and the actual observed life expectancy from the data. This most likely has to do with the varying factors among different countries including other health factors such as air quality, sanitation, living conditions, and political stability. More conclusive research would have to be done to verify this. Possible methods such as multivariate regression and an analysis of covariance could be done. It is known that a person's health is affected by many factors, and a complete study to analyze the effect of each factor would be extensive. Though a study of this scale is improbable for this paper, such findings could revolutionize what public health organizations, governments, and citizens prioritize in relation to their health. This type of study has far-reaching effects and could not only improve a person's life expectancy but also their quality of life.

\section{Conclusion}

Studies regarding drinking water have been done to draw conclusions about the presence of certain diseases when clean drinking water is limited. Research revolves around specific diseases or access to clean drinking water based on a scale. In some cases, the distance the water source is from the community is not measured. This research paper focuses on investigating the effect of the proportion of a population's access to basic drinking water services on their life expectancy. This research filters the kind of access to clean water by requiring availability to be within 30 minutes round trip as well as provides a more widespread conclusion than those regarding specific water-borne illnesses. We have found a positive correlation between the population's access to basic drinking water servies and life expectancy. This supports existing research that clean water is essential for the body to function, yet expands the relationship to overall life expectancy which is impacted by numerous factors within both physical and mental health. Insights in this field can further promote the basic right to safe water within a reasonable distance from one's home and highlight just how important water is in all aspects of human health.
